Smooth and Persistent Forecasts of German GDP\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}

### 2025-08-08 Achive IWH Simon paper/08.08.2025/paper.tex
up

\usepackage{graphicx} % Required for inserting images
\usepackage{tabularx}
\usepackage{natbib}
\usepackage[spaces,hyphens]{url}
\usepackage[hidelinks]{hyperref}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{hyphenat}
\usepackage{color}
%\usepackage{subfig}
% \usepackage[justification=left,hang]{caption}
\usepackage[justification=raggedrig,hang]{caption}
\usepackage{subcaption}
	%\captionsetup[subfigure]{labelfont=normal,textfont=normalfont,singlelinecheck=off,justification=centering}
\usepackage[table]{xcolor} 
\usepackage{colortbl}
\usepackage[hang]{footmisc}
\usepackage{appendix}
\usepackage{float}

@@

\title{Smooth and Persistent Forecasts of German GDP}
\author{Katja Heinisch\thanks{Halle Institute for Economic Research (IWH).} \hspace{0.2cm}
Simon van Norden\thanks{HEC Montréal and CIREQ.} \hspace{0.2cm}
Marc Wildi\thanks{Zurich University of Applied Sciences (ZHAW).}}
\date{May 2025 \\
\vspace{2cm} \textit{Work in progress, not for circulation}}

@@

\begin{abstract} %max.
This paper addresses the trade-off between informational efficiency and forecast smoothness in economic forecasting. Efficient forecasts are often volatile, limiting their practical usefulness. We propose a novel method that employs the framework of Multivariate Smooth Sign Accuracy (M-SSA) to extract smoothed components from leading indicators. This approach enhances the signal-to-noise ratio and allows for systematic control over forecast volatility and timing. Applied to German GDP, our method yields smoothed forecasts that improve the accuracy of traditional models, particularly over medium-term horizons. While smoother forecasts may lag slightly, this can be mitigated by adjusting the forecast horizon. Our findings offer practical tools for forecasters and policymakers subject to adjustment costs or frictions in updating forecasts or implementing policy changes.

@@

Economic forecasters face a dilemma: While efficiency suggests that their forecasts should incorporate all available information, the resulting forecasts are often volatile, which may reflect greater information content or contamination by excessive ``noise''. Further, forecast users (including public and private decision makers) may be constrained by how rapidly they can adjust to changing forecasts.\footnote{For example, private or public spending decisions may be constrained by annual budgets. Similarly, central banks increasingly attempt to smooth interest rate changes to allow them to signal the future direction of policy to markets, while private firms attempt to smooth changes in dividends.}

%%% Further recent literature
% \cite{lehmann2021predicting} show the performance of ifo busines cycles indicator for the nowcast (h=0) and forecast (h=1)

\cite{Wildi2024,Wildi2025} and \cite{McElroy2019,McElroy2020} propose novel methods that optimally smooth forecasts by controlling the expected frequency of sign changes. The resulting Smooth Sign Accuracy (SSA) framework encompasses efficient mean-square error (MSE) as a special case and produces increasingly smooth forecasts as changes in forecast sign are increasingly penalized.

@@

This study proposes a methodological innovation that extends the classical direct forecasting framework by integrating dimension-reduced information extracted through Multivariate Smooth Sign Accuracy (M-SSA). The core idea is to replace raw
%, unprocessed 
leading indicators with smoothed and filtered components derived from the M-SSA procedure, thereby improving the signal-to-noise ratio in the predictor variables used to forecast GDP. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP and GDP trends using a variety of sample periods and performance metrics.

Surprisingly, we find that smoothing indicator-based forecasts often provides useful leading information for GDP, especially over mid-term horizons ranging from two to four quarters ahead. We also show that while smoother GDP forecasts lag efficient MSE forecasts slightly, this can be mitigated by adjustments in the forecast horizon.\footnote{The lag reflects the trilemma highlighted by \cite{McElroy2019} in which forecasters choose between forecast accuracy, timeliness and volatility.}

%Previous research often highlights the utility of direct forecasts, particularly for short-term predictions such as nowcasting and forecasts one quarter ahead. However, this paper advances the literature by showing that more sophisticated predictors—specifically, filtered components obtained via M-SSA yield superior performance, especially over mid-term horizons ranging from two to four quarters ahead.

Section \ref{sec:data} describes our data series, the sample period used, and documents their dynamic behaviour. Section \ref{sec:direct_forecast} provides benchmark forecasts using simple direct projections of GDP based on our indicators. We demonstrate how in-sample forecast performance may improve when we ``de-noise'' the indicators prior to projection with simple smoothing filters. Section \ref{sec:mSSA} defines the M-SSA method and shows how its forecast performance compares to that of direct projections as well as how it varies with forecast horizon. Section \ref{sec:mSSA_component} considers the ability of our indicators to forecast German GDP growth using the M-SSA framework, and Section \ref{sec:revisions} investigates the temporal stability of the estimated forecasting equations. Section \ref{sec:conclusions} summarizes the results and concludes. The Appendix provides some additional results.

\section{Data and Dependence}\label{sec:data}
\subsection{Data}
The analysis employs a set of well-established leading indicators to forecast quarterly German GDP \citep{Drechsel2012financial,Heinisch2018bottom}, including industrial production (IP), the ifo Business Climate Index (ifo\_c), the Economic Sentiment Indicator (ESI), and the term spread between 10-year bond yields and the 3-month EURIBOR rate (spre\_10y\_3m) covering the period 1995 to 2024.  
We take into account the publication lag of the indicators and the GDP series.\footnote{Data revisions are not considered, given that previous studies \citep{Heinisch2019} have shown that their overall effect on forecast performance is minor.} 
While GDP is available with a one-quarter delay (published in mid of the next quarter), and IP is released with a one-month lag, business surveys and financial data are published without any lag, see Table \ref{tab:data_stucture} which presents the data structure as of the end of January 2025, just before the publication of the Flash estimate for the previous quarter.\footnote{Starting in the second quarter of 2020, the German Federal Statistical Office published for the first time the gross domestic product (GDP) 30 days after the end of the quarter.} All series are consequently re-aligned to correspond with the information available to forecasters, with GDP and IP variables shifted forward according to their respective lags. Table \ref{tab:data_stucture_aligned} displays the dataset as of the end of 2024, with GDP and IP variables temporally aligned at the sample endpoint. This configuration reflects the matrix of explanatory variables accessible in Q4-2024 for the purposes of nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications. In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & ip & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-07-01 &  & 90.700 & 85.800 & 92.500 & -1.300 \\ 
  2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
  2024-09-01 &902.571  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
  2024-10-01 &  & 90.800 & 84.100 & 90.700 & -1.200 \\ 
  2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
  2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}
\caption{Ragged end as of January 2025, before publication of the flash estimate.} 
\label{tab:data_stucture}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrc}
  \hline
 & GDP & ip & ifo\_c & ESI & spr\_10y\_3m &&Target (nowcast)\\ 
  \hline
  2024-07-01 & & 93.400 & 85.800 & 92.500 & -1.300&|& \\ 
  2024-08-01 & & 90.700&  84.900 & 90.900 & -1.400&| &\\ 
  2024-09-01 & 901.623 &93.200 & 83.700 & 89.800 & -1.400&| &902.571\\ 
  2024-10-01 &  &91.200 & 84.100 & 90.700 & -1.200&| &\\ 
  2024-11-01 &  & 90.800& 83.800 & 89.300 & -0.900&| &\\ 
  2024-12-01 &902.571   & 92.200  & 82.800 & 86.900 & -0.800&| &?\\ 
    \hline
\end{tabular}
\caption{Columns 1 through 5 constitute the matrix of explanatory variables available in Q4-2024 for the establishment of a GDP nowcast or forecast. All series are artificially aligned at the sample endpoint. Column 6 represents the target variable for the nowcast in Q4-2024.} 
\label{tab:data_stucture_aligned}
\end{table}

To ensure comparability and mitigate scale effects, all indicators (except the spread, which is differenced only) are log-differenced and standardized, the latter applied solely for ease of visual inspection. Additionally, extreme outliers associated with the COVID-19 pandemic are trimmed to prevent distortions in subsequent figures. 
 Monthly series are transformed to quarterly series and are displayed in Figure \ref{fig:data}, while Figure \ref{fig:data_lags} offers a close-up view of the financial crisis and the pandemic.\footnote{Monthly series are averaged to quarterly values.} The latter figure illustrates that the real-time GDP and industrial production series are synchronized at dips and peaks (coincident), whereas the other indicators tend to lead somewhat. The latter can be leveraged in a multivariate approach to enhance forecasting performance compared to univariate benchmarks.

@@

\subsection{Dependence}
To examine the dynamic interdependencies among the variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \ref{sec:mSSA} and \ref{sec:mSSA_component}.

Figures \ref{CCF} (full sample) and \ref{CCF_wc} (omitting the Pandemic period) show the sample cross-correlation functions (CCF's) between current and lagged GDP and the two survey indicators. Both figures show strong correlations that peak at leads relative to GDP. Results for IP showed no lead, while correlations for the term spread were weaker. The figures also show that the precise strengths of the correlations are influenced by the outliers associated with the COVID-19 pandemic. In the interest of model stability, we exclude data from Q4-2019 to Q4-2020 (pandemic) when modelling the dynamics for the multivariate filters developed below.

@@

We then estimated a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}.  This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package.  This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The figures \ref{ma_inv_BIP} and \ref{ma_inv_BIP_var3_en} show the implied responses of GDP to shocks to each indicator variable.\footnote{Shown are the MA coefficients for GDP from the vector moving-average (VMA) models obtained by inversion of the VAR(1) and VAR(3) representations respectively (i.e. the impulse-response functions, or IRFs, of GDP to the correlated reduced-form shocks.)}

@@

%The models encapsulate key characteristics of the data after adjusting for publication lags. Notably, 
GDP exhibits the strongest response to shocks in the two survey indicators—ifo and ESI\footnote{There is also weaker evidence of a response to the term spread at longer lags.}. The impact of the pandemic modifies the lag structure by emphasizing shorter dependencies triggered by the alternating outliers during that specific episode. Removing the COVID-19 breakout from the analysis results in smoother, somewhat longer-tailed dependencies, consistent with the observed empirical autocorrelation functions. In particular, the VAR(3) model captures slightly extended lead times and more nuanced lag patterns\footnote{The VAR(3) model yields marginal improvements in multivariate filtering performance. However, due to its relative simplicity, this study primarily employs the VAR(1) model. The M-SSA package (accessible at \url{https://github.com/wiaidp/R-package-SSA-Predictor}) offers more advanced capabilities, including the implementation of the VAR(3) model and various additional (Bayesian) VAR specifications.}. The delayed response of GDP to movements in these indicators suggests that multivariate models could yield superior forecasting performance compared to univariate approaches, a hypothesis we explore in the next section

@@

In this section, we demonstrate how pre-filtering indicator variables may improve the performance of simple forecasts by removing noise. For simplicity, all the models in this section are direct projections of GDP on our indicators and take the form

\begin{equation}
    u_{t+h} = \beta' \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $u_{t+h}$ is the growth rate of GDP, $h = \{0, 1, 2, ... \}$
    \item $\mathbf{X}_t \equiv [1,u_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $u_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\beta}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

We then demonstrate how in-sample forecast performance improves when we ``de-noise'' the indicators prior to projection with simple smoothing filters.\footnote{The smoothing filters are one-sided (concurrent) filters of length 31, therefore, estimation of both forecasting models omits the first 30 observations.}

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_all.pdf}
        %\includegraphics[height=3in, width=4.5in]{./Figures/direct_wc_all.pdf}
        \caption{Direct forecasts (red lines) based on all indicators and full sample information (without pandemic): GDP (black line) left-shifted  by zero, one, two, and three quarters.
        \label{direct_wc}}
    \end{center}
\end{figure}

Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} compare actual and fitted values for GDP at forecast horizons $h = 0, \ldots, 3$, using the four indicators and the target, i.e. GDP shifted forward (leftwards) by $h$ quarters, without applying any filtering. Figure \ref{direct_wc} shows results over the sample without the pandemic, while Figure \ref{direct_wc_financial_crisis} provides a closer look at the fit during the financial crisis. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients on the indicators are zero (HAC-adjusted Wald test).

@@

The results indicate that as the forecast horizon $h$ increases, the turning points (i.e., peaks and troughs) of the direct forecasts become increasingly right-shifted (lagged) relative to the target GDP. Additionally, the p-values of the HAC-adjusted Wald test precipitously increase from $h=2$ to $h=3$ (see Table \ref{tab:pvaluedhp}), implying that the indicators become statistically insignificant for forecast horizons of more than two quarters.\footnote{ 
These findings are confirmed by the relative root mean-square errors comparing direct forecasts against the mean benchmark, see Table \ref{rRMSE_mSSA_direct_mean_without_covid8} in the Appendix.}

@@

The Hodrick-Prescott (HP) filter is a classic tool used in business cycle analysis: the filter is specified by a single smoothing parameter $\lambda$ and the value $\lambda=1600$ is specifically recommended for quarterly data. However, \cite{Phillips_Jin_2021} noted that the HP(1600) filter removes relevant information due to excessive smoothing. In our context, this oversmoothing issue would be further exacerbated when considering forecast horizons shorter than a year---an interval inconsistent with the mean duration of up to several years of business cycles, as highlighted by the HP(1600). To address this concern, we suggest a more adaptive HP(160) `target' filter.

Figure \ref{hp_160} presents both classic two-sided and one-sided HP(160) (denoted HP-C for \textit{concurrent}) filters (top panels) along with their associated amplitude functions in the frequency domain (bottom panels). It confirms that the HP-C filter assigns greater weight to yearly components—relevant in a mid-term forecast context— than the classic quarterly HP filter, while also reducing undesirable high-frequency noise. Therefore, while the choice of the HP(160) filter in this example
is somewhat arbitrary, it should serve the purpose of reducing high-frequency noise while preserving longer-duration movements.\footnote{A comprehensive technical analysis of the effect of $\lambda$ on the resulting GDP predictor can be conducted using the M-SSA package (\url{https://github.com/wiaidp/R-package-SSA-Predictor}).}

\begin{figure}[h]
    \begin{center}
        \includegraphics[height=4in, width=5.5in]{./Figures/hp_160.pdf}
        \caption{Two-sided HP(160) (top left) and one-sided (concurrent) HP-C filters (top right) with corresponding amplitude functions (bottom).
        \label{hp_160}}
    \end{center}
\end{figure}

>>>>

\usepackage{graphicx} % Required for inserting images
\usepackage{tabularx}
\usepackage{natbib}
\usepackage[spaces,hyphens]{url}
\usepackage[hidelinks]{hyperref}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{hyphenat}
\usepackage{color}
%\usepackage{subfig}
% \usepackage[justification=left,hang]{caption}
\usepackage[justification=raggedright,hang]{caption}
\usepackage{subcaption}
	%\captionsetup[subfigure]{labelfont=normal,textfont=normalfont,singlelinecheck=off,justification=centering}
\usepackage[table]{xcolor} 
\usepackage{colortbl}
\usepackage[hang]{footmisc}
\usepackage{appendix}
\usepackage{float}

@@

\title{Smooth and Persistent Forecasts of German GDP\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}}
\author{Katja Heinisch\thanks{Halle Institute for Economic Research (IWH).} \hspace{0.2cm}
Simon van Norden\thanks{HEC Montréal and CIREQ.} \hspace{0.2cm}
Marc Wildi\thanks{Zurich University of Applied Sciences (ZHAW).}}
\date{July 2025 \\
\vspace{2cm} \textit{Work in progress, not for circulation}}

@@

\begin{abstract} %max.
This paper explores the trade-off between informational efficiency and forecast smoothness in economic forecasting. Forecasts that minimize mean squared forecast errors often exhibit excessive volatility, limiting their practical applicability. We propose the Multivariate Smooth Sign Accuracy (M-SSA) framework as a solution, extracting smoothed components from leading indicators. This enhances the signal-to-noise ratio and controls forecast volatility and timing. Applied to German GDP, our method yields smoothed forecasts that improve the forecasting accuracy of traditional models, particularly over medium-term horizons. While smoother forecasts tend to lag slightly, this drawback can be mitigated by adjusting the forecast horizon. These findings highlight the practicality of the M-SSA framework for both forecasters and policymakers, especially in settings where updating forecasts or policy adjustments are costly.

@@

%Economic forecasters face a dilemma: While efficiency suggests that their forecasts should incorporate all available information, the resulting forecasts are often volatile, which may reflect greater information content or contamination by excessive ``noise''. Further, forecast users (including public and private decision makers) may be constrained by how rapidly they can adjust to changing forecasts.\footnote{For example, private or public spending decisions may be constrained by annual budgets. Similarly, central banks increasingly attempt to smooth interest rate changes to allow them to signal the future direction of policy to markets, while private firms attempt to smooth changes in dividends.}

Economic forecasters face a dilemma
: while informational efficiency requires that forecasts incorporate all available data, doing so often results in forecasts that are excessively volatile.
 %Volatile forecasts may be a sign that they efficiently incorporate all available information. 
 In addition, forecasts may reflect excessive ``noise'' due to overfitting. Further, forecast users (including public and private decision makers) may face constraints in their speed of adjustment to changing forecasts. For example, fixed annual budgets may constrain spending, or transaction costs may limit changes in investments. Others may attempt to smooth their reactions for reasons of signalling. Important examples include dividend smoothing by corporate managers and interest rate smoothing by central banks.

\cite{Wildi2024,Wildi2025} and \cite{McElroy2019,McElroy2020} propose novel methods that optimally smooth forecasts by controlling the expected frequency of sign changes. The resulting Smooth Sign Accuracy (SSA) framework encompasses efficiently the mean-square error (MSE) as a special case and produces increasingly smooth forecasts as changes in the forecast direction (sign) are increasingly penalized.

@@

%This study proposes a methodological innovation that extends the classical direct forecasting framework by integrating dimension-reduced information extracted through Multivariate Smooth Sign Accuracy (M-SSA). The core idea is to replace raw
%, unprocessed 
%leading indicators with smoothed and filtered components derived from the M-SSA procedure, thereby improving the signal-to-noise ratio in the predictor variables used to forecast GDP. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP and GDP trends using a variety of sample periods and performance metrics.

This study extends this framework by replacing raw leading indicators with smoothed and filtered components, thereby improving the signal-to-noise ratio in the predictor variables. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP \citep{lehmann2021predicting,Heinisch2018bottom}  and GDP trends using a variety of sample periods and performance metrics.

We find that smoothing indicator-based forecasts often provides useful leading information for GDP, especially over medium-term horizons ranging from two to four quarters ahead. We also show that while smoother GDP forecasts lag efficient MSE forecasts slightly, this can be mitigated by adjustments in the forecast horizon.\footnote{The lag reflects the trilemma highlighted by \cite{McElroy2019} in which forecasters choose between forecast accuracy, timeliness, and volatility.}

%Previous research often highlights the utility of direct forecasts, particularly for short-term predictions such as nowcasting and forecasts one quarter ahead. However, this paper advances the literature by showing that more sophisticated predictors—specifically, filtered components obtained via M-SSA yield superior performance, especially over medium-term horizons ranging from two to four quarters ahead.

We proceed as follows. Section \ref{sec:data} describes our data series, the sample period used, and documents their dynamic behaviour. Section \ref{sec:direct_forecast} provides benchmark forecasts using simple direct projections of GDP based on selected indicators. We demonstrate how in-sample forecast performance may improve when we ``de-noise'' the indicators prior to projection with simple smoothing filters. Section \ref{sec:mSSA} defines the M-SSA method and shows how its forecast performance compares to that of direct projections as well as how it varies with forecast horizon. 
Section \ref{sec:mSSA_component} considers the ability of our indicators to forecast German GDP growth out-of-sample using the M-SSA framework, and Section \ref{sec:revisions} investigates the temporal stability of the estimated forecasting equations. 
Section \ref{sec:conclusions} summarizes our results and concludes.

\section{Data and Dependence}\label{sec:data}
\subsection{Data and Publication Lags}
The analysis employs a set of well-established leading indicators to forecast quarterly German GDP \citep{Drechsel2012financial,Heinisch2018bottom}, including industrial production (IP), the ifo Business Climate Index (ifo\_c), the Economic Sentiment Indicator (ESI), and the term spread between 10-year bond yields and the 3-month EURIBOR rate (spre\_10y\_3m) covering the period 1995 to 2024.  
We take into account the publication lag of the indicators and the GDP series (Table \ref{tab:data_stucture}).\footnote{Data revisions are not considered, given that previous studies \citep{Heinisch2019} have shown that their overall effect on forecast performance is minor.} 
While GDP is available with a one-quarter delay, and IP is released with a two-month lag, business surveys and financial data are published without any lag. %All series are consequently re-aligned to correspond with the information available to forecasters, with GDP and IP variables shifted forward according to their respective lags. Table \ref{tab:data_stucture_aligned} displays the dataset as of the end of 2024, with GDP and IP variables temporally aligned at the sample endpoint. This configuration reflects the matrix of explanatory variables accessible in Q4-2024 for the purposes of nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications. In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-07-01 &  & 90.700 & 85.800 & 92.500 & -1.300 \\ 
  2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
  2024-09-01 &902.571  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
  2024-10-01 &  & 90.800 & 84.100 & 90.700 & -1.200 \\ 
  2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
  2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}
\caption{Ragged edge structure. As of January 2025, before publication of the GDP flash estimate.} 
\label{tab:data_stucture}
\end{table}

To allow for missing observations due to publication lags, we re-align series to reflect the information available to forecasters. Table \ref{tab:data_stucture_aligned} displays the result as of the end of 2024, with GDP and IP columns shifted to reflect their release dates rather than the periods that they measure. This will be the data configuration used for nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications.\footnote{In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.}

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrc}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m &&Target (nowcast)\\ 
  \hline
  2024-07-01 & & 93.400 & 85.800 & 92.500 & -1.300&|& \\ 
  2024-08-01 & & 90.700&  84.900 & 90.900 & -1.400&| &\\ 
  2024-09-01 & 901.623 &93.200 & 83.700 & 89.800 & -1.400&| &902.571\\ 
  2024-10-01 &  &91.200 & 84.100 & 90.700 & -1.200&| &\\ 
  2024-11-01 &  & 90.800& 83.800 & 89.300 & -0.900&| &\\ 
  2024-12-01 &902.571   & 92.200  & 82.800 & 86.900 & -0.800&| &?\\ 
    \hline
\end{tabular}
\caption{Aligned dataset. At each date, columns 1 through 5 show the most recent values of each predictor available to forecasters. Column 6 represents the target variable for a GDP nowcast (i.e. the values of column 1 shifted upwards by the publication lag of GDP.} 
\label{tab:data_stucture_aligned}
\end{table}

To ensure comparability and mitigate scale effects, all indicators (except the term spread, which is differenced only) are log-differenced and standardized. Additionally, extreme outliers associated with the COVID-19 pandemic are trimmed to prevent distortions in subsequent figures. 
 Monthly series are shifted (Table \ref{tab:data_stucture_aligned}) and averaged to quarterly series (Figure \ref{fig:data}), while Figure \ref{fig:data_lags} offers a close-up view of the financial crisis and the pandemic. The latter figure shows that the real-time GDP and industrial production series are synchronized at turning points, whereas the other indicators tend to lead somewhat. The challenge for forecasters is to use the latter in a multivariate approach to improve forecasting performance over that of univariate benchmarks.

@@

\subsection{Dependence}
To examine the joint dynamics of our variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \ref{sec:mSSA} and \ref{sec:mSSA_component}.

Figures \ref{CCF} (full sample) and \ref{CCF_wc} (omitting the Pandemic period) show the sample cross-correlation functions (CCF's) between current and lagged GDP and the two survey indicators. Both figures show strong correlations that peak when GDP is slightly lagged. Results for IP showed no such lag, while correlations for the term spread were weaker. The figures also show that the precise strengths of the correlations are influenced by the outliers associated with the COVID-19 pandemic. In the interest of model stability, we exclude data from Q4-2019 to Q4-2020 (pandemic) when modelling the dynamics for the multivariate filters developed below.

@@

We then estimated a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}.  This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package.  This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The figures \ref{ma_inv_BIP} and \ref{ma_inv_BIP_var3_en} show the implied responses of GDP to shocks to each indicator variable.\footnote{Shown are the MA coefficients for GDP from the vector moving-average (VMA) models obtained by inversion of the VAR(1) and VAR(3) representations respectively (i.e. the impulse-response functions (IRFs) of GDP to the correlated reduced-form shocks.)}

@@

%The models encapsulate key characteristics of the data after adjusting for publication lags. Notably, 
GDP exhibits the strongest response to shocks in the two survey indicators—ifo and ESI.\footnote{There is also weaker evidence of a response to the term spread at longer lags.} The impact of the pandemic modifies the lag structure by emphasizing shorter dependencies triggered by the alternating outliers during that specific episode. Removing the pandemic from the analysis results in smoother, somewhat longer-tailed dependencies, consistent with the observed empirical autocorrelation functions. In particular, the VAR(3) model captures slightly longer lead times and more nuanced lag patterns.\footnote{The VAR(3) model yields marginal improvements in multivariate filtering performance. However, due to its relative simplicity, this study primarily employs the VAR(1) model. The M-SSA package (accessible at \url{https://github.com/wiaidp/R-package-SSA-Predictor}) offers more advanced capabilities, including the implementation of the VAR(3) model and Bayesian VAR specifications.}

The delayed response of GDP to movements in the survey indicators suggests that multivariate models could yield better forecasts than those of univariate model, a hypothesis we explore in the next section.

@@

In this section, we demonstrate how pre-filtering indicator variables may improve the performance of simple forecasts by removing noise. We then demonstrate how in-sample forecast performance improves when we ``de-noise'' the indicators prior to projection with simple smoothing filters.\footnote{The smoothing filters are one-sided (concurrent) filters of length 31, therefore, estimation of both forecasting models omits the first 30 observations.}

For simplicity, all the models in this section are direct projections of GDP on our indicators and take the form

\begin{equation}
    u_{t+h} = \boldsymbol{\beta'} \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $u_{t+h}$ is the quarterly growth rate of GDP for forecast horizon $h = \{0, 1, 2, ... \}$
    \item $\mathbf{X}_t \equiv [1,u_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $u_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\boldsymbol{\beta}}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_all.pdf}
        %\includegraphics[height=3in, width=4.5in]{./Figures/direct_wc_all.pdf}
        \caption{Direct forecasts (red lines) based on all indicators and full sample information (without pandemic.)\\
        Target (black line): GDP left-shifted  by 0 - 3 quarters.
        \label{direct_wc}}
    \end{center}
\end{figure}

Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} compare actual and fitted values for GDP at forecast horizons $h = 0, \ldots, 3$, using the four indicators and the target, i.e., GDP shifted $h$ quarters ahead. Figure \ref{direct_wc} shows results over the sample without the pandemic, while Figure \ref{direct_wc_financial_crisis} provides a closer look at the fit during the financial crisis. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients on the indicators are zero (HAC-adjusted Wald test).

@@

The results indicate that as the forecast horizon $h$ increases, forecast become less volatile and their turning points (i.e., peaks and troughs) become increasingly right-shifted (lagged) relative to the target GDP. Additionally, the p-values of the HAC-adjusted Wald test precipitously increase from $h=2$ to $h=3$ (see Table \ref{tab:pvaluedhp}), implying that the indicators become statistically insignificant for forecast horizons of more than two quarters.\footnote{ 
These findings are confirmed by the relative root mean-square errors comparing direct forecasts against the mean benchmark, see Table \ref{rRMSE_mSSA_direct_mean_without_covid8} in the Appendix.}

%\begin{table}[ht]
%\centering
%\begin{tabular}{rrrrr}
%  \hline
% & h=3 & h=4 & h=5 & h=6 \\ 
%  \hline
%rRMSE & 0.962 & 0.967 & 0.990 & 0.967 \\ 
%   \hline
%\end{tabular}
%\caption{Relative root mean-square error of HP-C against classic direct forecast for $h\geq %3$.\\Values $<1$ imply  HP-C has smaller RMSE.  } 
%\label{tab:rRMSE}
%\end{table}
\bigskip
\begin{table}[h]
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}
\caption{p-values for $H_0: \boldsymbol{\beta} = 0$ in \ref{eq:direct_forecast} based on HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\end{table}

@@

The Hodrick-Prescott (HP) filter is a classic tool used in business cycle analysis: the filter is specified by a single smoothing parameter $\lambda$. The value $\lambda=1600$ is the most common value used for business cycle analysis, but \cite{Phillips_Jin_2021} note that this removes relevant information due to excessive smoothing. 
%In our context, this oversmoothing issue would be further exacerbated when considering forecast horizons shorter than a year---an interval inconsistent with the mean duration of up to several years of business cycles, as highlighted by the HP(1600). 
To address this concern, we instead use $\lambda = 160$ to allow for a more adaptive filter.

Figure \ref{hp_160} presents both classic two-sided and one-sided HP(160) (denoted HP-C for \textit{concurrent}) filters (top panels) along with their associated amplitude functions in the frequency domain (bottom panels). It confirms that the HP-C filter assigns greater weight to yearly components—relevant in a medium-term forecast context— than the classic quarterly HP filter, while also reducing undesirable high-frequency noise. Therefore, while the choice of the HP(160) filter in this example
is somewhat arbitrary, it should serve the purpose of reducing high-frequency noise while preserving longer-duration movements.\footnote{A comprehensive technical analysis of the effect of $\lambda$ on the resulting GDP predictor can be conducted using the M-SSA package (\url{https://github.com/wiaidp/R-package-SSA-Predictor}).}
